\documentclass[preprint]{sigplanconf}

\usepackage{amsmath}
\usepackage{color}
\usepackage{graphicx}
\usepackage[nice]{nicefrac}

\input{src/wgp/defs}

\begin{document}

\conferenceinfo{SIGPLAN WGP '11}{Sept. 9, 2011, Tokyo, Japan.} 
\copyrightyear{2011}
\copyrightdata{[to be supplied]} 

\titlebanner{banner above paper title (pre-print only)}        % These are ignored unless
\preprintfooter{short description of paper (pre-print only)}   % 'preprint' option specified.

\title{[a better title]}
%\subtitle{Subtitle Text, if any}

\authorinfo{Moss Prescott and Jeremy Siek}
           {University of Colorado, Boulder}
           {\{moss.prescott,jeremy.siek\}@colorado.edu}

\maketitle

\begin{abstract}
% Shooting for ~200 words:
%\input{src/abstract}
\todo{Tried to start here but it was a disaster so instead worked on reducing the thesis to a compressed outline instead.}
\end{abstract}

\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}  % ???

\terms
Languages, Syntax

\keywords
Syntax extension, Domain Specific Languages, Abstract syntax


\section{new outline}

intro: position rel. to other work. summarize entire paper.; the most important connected ideas

define what the thing is, what the methodology is (lang. oriented meta-prog?)

thesis:

- methodology enables system to be an order of mag. smaller than same thing using previous approaches

-- wider range of extensions

-- also simpler/smaller

-- prototype

-- case studies that illustrate


- lessons learned




section 2: case studies

section 3: overview of system; what are all the pieces

section 4: ASTs and Txs; kernel language

section 5: presentation language

section 6: related work

section 7: conclusion


\input{src/wgp/intro}

\section{Two (or three) approaches}
\subsection{Text-based}
Familiar and well-understood. Difficulties inherent in parsing; limitations of text.

\subsection{Structure Editors and Language Workbenches}
Successful for certain domains. Inflexibility, large footprint, etc.

\subsection{LISP}
Eschew concrete syntax (but not really completely).

\input{src/wgp/ast}

\input{src/wgp/tx}

\input{src/wgp/studies}

\section{Related Work}
[Not the language workbenches, but other AST-based work?]

\section{Conclusion}
~

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

%\acks

%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}
