\chapter{Objective Symptoms}
% \OnePageChapter	% (use if this chapter/appendix is 1 page long)

\paragraph{Appendices} follow the same page-numbering rules
	as regular chapters.
	The first page of a multi-page appendix is not numbered.
	But the page of a single-page appendix {\em is} numbered.

\textbf{Are they slow learners} or is it a \emph{REAL} problem?
These are classic findings in the hopelessly computer challenged.

\begin{enumerate}

\item  Can't copy from hard drive to disk.
\item  Can't eject disks.
\item  The word ``disk'' has thousands of meanings to them. None are correct.
\item  Saving a document in any form is a concept totally unexplainable to them.
\item  Desktop covered with Untitled Folders - look again,
		untitled folders are everywhere.
\item  ``Lost'' documents found often in the Apple Menu.
\item  Trash always full. Claim they don't know how to place things in trash.
\item  Mysterious things happen to their documents or computer
	when they are not present.  AKA ``computer victims''.
\item  Highlighting = deleting. Dragging = Oblivion.
\item  Selecting, double-clicking a problem?
		They will always say their mouse is broken.
\item  Their double- click mechanics wants you to send them to a neurologist.
\item  Computer always on due to fear of having to restart it.
\item  Have never read their QuickMail - will say ``I prefer a phone call''.
\item  Have magical beliefs about what computers do.
\item  Describes some flaky way computers could REALLY help them,
		but is not yet available.
\item  Constantly saying they need more ``memory''.
\item  Requests gizmos and gadgets, i.e., ``mouse leash'' or ``disk cozy''.
\item  Avoids eye contact when talking about computers.

\end{enumerate}
